\section{Introduction}

\begin{frame}{Introduction}
    Sorting algorithms play a crucial role in computer science as they efficiently organize and process data.\cite{1}
    This paper focuses on the comparative analysis of two sorting algorithms: insertion sort and quick sort, implemented in three popular programming languages - C++, Java, and Python.
\end{frame}

\begin{frame}{Introduction (Contd.)}
    \begin{itemize}
        \item Our main objective is to evaluate the performance of these algorithms by measuring their execution time and identifying factors that influence their efficiency across different programming environments.
        \item Insertion sort was developed using iterative programming approaches, while quick sort was implemented utilizing recursive programming approaches.
        \item \cite{1}\cite{2}
        The programming languages were categorized into groups based on their characteristics:
            \begin{itemize}
                \item Virtual machine-based languages: Java (compiled and interpreted language)
                \item Native Languages: C++ (Compiled language)
                \item Scripting Languages: Python (Interpreted language)\cite{3}
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Introduction (Contd.)}
    \textbf{Key Goals:}
    \begin{itemize}
        \item Implementing both algorithms in C++, Java, and Python for a basis of comparison.
        \item Conducting a performance analysis by measuring execution times for varying input array sizes.
        \item Investigating the impact of programming language and implementation strategies on algorithm performance.
        \item Identifying scenarios where one algorithm may outperform the other based on data characteristics and programming environments.
        \item Providing insights and recommendations for developers to make informed decisions when selecting and implementing sorting algorithms in their applications.
    \end{itemize}
\end{frame}

\begin{frame}{Introduction (Contd.)}
    \textbf{Anticipated Contributions:}
    \begin{itemize}
        \item Illuminate the strengths and weaknesses of insertion sort and quick sort across programming languages.
        \item Offer guidance to developers on optimizing sorting algorithms based on use cases and performance requirements.
        \item Contribute to an understanding of algorithm design and implementation choices for data management.\\
        \cite{2}\cite{3}\cite{4}
    \end{itemize}
\end{frame}
